GVN-Hoist: Hoisting Computations from Branches

### 2016-08-25 hoist.tex
add more detailed description of GCC hoisting from Danny, and added an Ack section

There are a lot of bug reports in GCC and LLVM bugzillas
\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current LLVM implementation of code
hoisting in SimplifyCFG.cpp $HoistThenElseCodeToIf()$ is very limited to
hoisting from identical basic blocks: the instructions of two sibling basic
blocks are read in the same time, and all the instructions of the blocks are
hoisted to the common parent block as long as the compiler is able to prove that
the instructions are equivalent.  This implementation does not allow for an easy
extension: first in terms of compilation time overhead the implementation would
become quadratic in number of instructions to bisimulate, and second the
equivalence of instructions currently uses $$I1->isIdenticalToWhenDefined(I2)$$
would need to be rewritten to be more general, leading to using a mechanism
similar to GVN.

@@

Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions are
computed which are hoisted in a basic block where the expression is anticipable
(all the operands are available). This algorithm would hoist as far as possible
without regarding the impact on register pressure and as such a cost model will
be required.

GCC recently got code-hoisting which is implemented as part of GVN-PRE. The
algorithm hoists top down in a predecessor where the value is in ANTIC\_IN. This
approach may have the problem of hoisting the expression too far and to mititage
that it only hoists expressions when at least one successor computes it. Hence
the algorithm will hoist a limited set of computations.  One advantage of GCC
implementation is that it works in sync with the GVN-PRE such that when new
hoisting opportunities are created by GVN-PRE, code-hoisting will hoist them.

@@

>>>>

There are a lot of bug reports in GCC and LLVM bugzillas
\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current LLVM implementation of code
hoisting in SimplifyCFG.cpp $HoistThenElseCodeToIf()$ is very limited to
hoisting from identical basic blocks: the instructions of two sibling basic
blocks are read in the same time, and all the instructions of the blocks are
hoisted to the common parent block as long as the compiler is able to prove that
the instructions are equivalent.  This implementation does not allow for an easy
extension: first in terms of compilation time overhead the implementation would
become quadratic in number of instructions to bisimulate, and second the
equivalence of instructions currently uses $$I1->isIdenticalToWhenDefined(I2)$$
would need to be rewritten to be more general, leading to using a mechanism
similar to the idea described in this paper based on GVN.

@@

Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
are computed which are hoisted in a basic block where the expression is
anticipable (all the operands are available). This algorithm would hoist as far
as possible without regarding the impact on register pressure and as such a cost
model will be required.  Also the description of VBE is based on the classic
dataflow model and an adaptation to a sparse SSA representation is required.

GCC recently got code-hoisting which is implemented as part of GVN-PRE: it uses
the set of ANTIC and AVAIL value expressions computed for PRE.  The algorithm
hoists top down in a predecessor where the value is in ANTIC\_IN.  It uses
ANTIC[B] to know what expressions will be computed on every path from B to exit,
and can be computed in B: that is the safety condition.  It uses AVAIL[B] to
subtract out those values already being computed (because they are already
available): this is a redundancy condition.  The cost function is: for each
hoist candidate, if all successors of B are dominated by B, then we know
insertion into B will eliminate all the remaining computations.  It then checks
to see if at least one successor of B has the value available.  This avoids
hoisting it way up the chain to ANTIC.  It also checks to ensure that B has
multiple successors, since hoisting in a straight line is pointless.  The
algorithm continues on down the dominator tree, iterating with PRE until no more
changes.  One advantage of GCC implementation is that it works in sync with the
GVN-PRE such that when new hoisting opportunities are created by GVN-PRE,
code-hoisting will hoist them.

@@

\section{Acknowledgments}
We would like to thank Daniel Berlin for his code reviews and for his feedback
on earlier versions of this paper.
